---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  out.width = "80%"
)
```

# celltalker

The goal of celltalker is to infer putative ligand and receptor interactions from single-cell RNAseq data. This is accomplished by evaluating known cognate ligand/receptor interactions across groups of cells. Interactions are scored by jointly weighting the expression levels of ligands and receptors, and significance is evaluated by comparing to a background distribution of scrambled ligands and receptors.

A recent refactoring of this package has made the interface much simpler to use. We also provide a wrapper to the functionality of the [circlize package](https://jokergoo.github.io/circlize_book/book/) for creating circos plots of ligands and receptors.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("arc85/celltalker")
```

## TL;DR example use

Here's a basic use case for celltalker based on a subset of the 40,000 bone marrow single-cell dataset available from the Human Cell Atlas. Check out SeuratData [here](https://github.com/satijalab/seurat-data) to download the example dataset.

Note that we have identified canonical immune cell types and filtered the dataset for ease of analysis. A vignette documenting this processing is coming soon, but please utilize the **hca_bm_umap_cell_types** data.frame for the vignette below.

```{r example}
# Load packages
suppressMessages({
  library(celltalker)
  library(Seurat)
  suppressWarnings(
    library(SeuratData)
  )
  library(tidyverse)
})

# Load Human Cell Atlast Bone Marrow from SeuratData
data(hcabm40k)

# Filter cell and assign cell types to the dataset
# NB: hca_bm_umap_cell_types has cell types and UMAP embeddings
hca_bm <- hcabm40k[,rownames(hca_bm_umap_cell_types)]
hca_bm[["cell_types"]] <- hca_bm_umap_cell_types$cell_types

# Process data
hca_bm <- NormalizeData(hca_bm)

# Add UMAP coordinates
hca_bm[["umap"]] <- CreateDimReducObject(embeddings=as.matrix(hca_bm_umap_cell_types[,1:2]),
  key="UMAP_",assay="RNA")

# View cell types
DimPlot(hca_bm,group.by="cell_types")

## Run celltalker
hca_bm_interactions <- celltalk(input_object=hca_bm,
  metadata_grouping="cell_types",
  ligand_receptor_pairs=ramilowski_pairs,
  number_cells_required=100,
  min_expression=1000,
  max_expression=20000,
  scramble_times=10)

## Identify top statistically significant interactions
top_stats <- hca_bm_interactions %>%
  mutate(fdr=p.adjust(p_val,method="fdr")) %>%
  filter(fdr<0.05) %>%
  group_by(cell_type1) %>%
  top_n(3,interact_ratio) %>%
  ungroup()

## Generate a circos plot
colors_use <- RColorBrewer::brewer.pal(n=length(unique(hca_bm$cell_types)),"Set2")

circos_plot(ligand_receptor_frame=top_stats,
  cell_group_colors=colors_use,
  ligand_color="blue",
  receptor_color="red",
  cex_outer=0.5,
  cex_inner=0.4)

```
