---
title: "Identify significantly different interactions across groups with replicate samples"
author: "Anthony R Cillo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identify significantly different interactions across groups with replicate samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates an extension of the core functionality of celltalker to replicate samples. The idea is that this functionality can be used to identify consistently differentially expressed ligand and receptor interactions. To demonstrate this functionality, we will leverage our [previously published](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE139324) dataset of healthy donor PBMC and healthy donor tonsils.

The data preprocessing steps are documented in the data-raw directory. We will start with a low-dimensional visualization with annotated cell types.

## Load packages and setup analysis

```{r setup}
library(celltalker)
library(Seurat)
library(tidyverse)

data("filtered_lig_rec")
data("overall_metadata")
data("overall_umap")

## Create a Seurat object based on 5 healthy donor PBMC and 5 healhy donor tonsils
ser_obj <- CreateSeuratObject(filtered_lig_rec,meta.data=overall_metadata)
ser_obj[["umap"]] <- CreateDimReducObject(embeddings = overall_umap, key = "UMAP_", assay = DefaultAssay(ser_obj))

DimPlot(ser_obj,group.by="cell_types",split.by="sample_type")

```

## Run celltalker

Next, we will split the data into PBMC and Tonsil Seurat objects and will identify the top ligand and receptor interactions in each of these datasets

```{r}

## Split dataset
ser_split <- SplitObject(ser_obj,split="sample_type")

## Check out the split data
ser_split

## Run celltalker - PBMC
pbmc_interactions <- celltalk(input_object=ser_split[["PBMC"]],
  metadata_grouping="cell_types",
  ligand_receptor_pairs=ramilowski_pairs,
  number_cells_required=50,
  min_expression=1000,
  max_expression=20000,
  scramble_times=10)

## Check out the interactions - PBMC
pbmc_interactions %>%
  mutate(p_val_adj=p.adjust(p_val,method="fdr")) %>%
  filter(p_val_adj<0.05)

## Run celltalker - Tonsil
tonsil_interactions <- celltalk(input_object=ser_split[["Tonsil"]],
  metadata_grouping="cell_types",
  ligand_receptor_pairs=ramilowski_pairs,
  number_cells_required=50,
  min_expression=1000,
  max_expression=20000,
  scramble_times=10)

## Check out the interactions - tonsil
tonsil_interactions %>%
  mutate(p_val_adj=p.adjust(p_val,method="fdr")) %>%
  filter(p_val_adj<0.05)

```

## Create circos plots

```{r}

## Plot top 3 interactions - PBMC
top_stats_pbmc <- pbmc_interactions %>%
  mutate(p_val_adj=p.adjust(p_val,method="fdr")) %>%
  filter(p_val_adj<0.05) %>%
  group_by(cell_type1) %>%
  top_n(3,interact_ratio) %>%
  ungroup()

# Assign colors to cell types
all_cell_types <- unique(ser_obj[["cell_types"]][,1])

colors_use <- colorRampPalette(RColorBrewer::brewer.pal(n=8,"Set2"))(length(all_cell_types))
names(colors_use) <- all_cell_types

suppressMessages(
circos_plot(ligand_receptor_frame=top_stats_pbmc,
  cell_group_colors=colors_use,
  ligand_color="blue",
  receptor_color="red",
  cex_outer=0.5,
  cex_inner=0.4)
)

## Plot top 3 interactions - Tonsil
top_stats_tonsil <- tonsil_interactions %>%
  mutate(p_val_adj=p.adjust(p_val,method="fdr")) %>%
  filter(p_val_adj<0.05) %>%
  group_by(cell_type1) %>%
  top_n(3,interact_ratio) %>%
  ungroup()

suppressMessages(
circos_plot(ligand_receptor_frame=top_stats_tonsil,
  cell_group_colors=colors_use,
  ligand_color="blue",
  receptor_color="red",
  cex_outer=0.5,
  cex_inner=0.4)
)

```

## Identify significantly different interactions across patient samples

```{r}

# Split Seurat object into lists
ser_obj_list <- SplitObject(ser_obj,split="sample_type")

# Evaluate significance of interactions across replicates
consistent_tonsil_interactions <-  compare_group_interactions(
  interactions_stats=top_stats_tonsil,
  sample_replicates="sample_id",
  sample_group_1=ser_obj_list[["Tonsil"]],
  sample_group_2=ser_obj_list[["PBMC"]])

# Significant interactions with type I error control with FDR
consistent_tonsil_interactions %>%
  mutate(p_val_adj=p.adjust(p_value,method="fdr")) %>%
  filter(p_val_adj<0.05)

# Boxplot of joint weight interaction across replicates
boxplot_group_interaction(ligand="CD40LG",
  receptor="TRAF3",
  cell_type1="CD4 T cells",
  cell_type2="Plasmablasts",
  sample_replicates="sample_id",
  sample_group_1=ser_split[["Tonsil"]],
  sample_group_2=ser_split[["PBMC"]])

```
